Weighted Model Counting as a Special Case of Polyadic Measure Algebras

### 2020-03-07 paper.tex
A detailed example of how a probability is computed

\documentclass{article}
\usepackage[utf8]{inputenc}
\usepackage[UKenglish]{babel}
\usepackage[UKenglish]{isodate}
\usepackage{fullpage}
\usepackage{amsthm}
\usepackage{amsfonts}
\usepackage{amsmath}
\usepackage{mathtools}
\usepackage[capitalise]{cleveref}
\usepackage{bm}
\usepackage{booktabs}

@@

\title{Statistical Relational Models as Polyadic Measure Algebras}
\author{Paulius Dilkas}

@@

% TODO: Which definition do I actually need?
\begin{definition}[\cite{DBLP:books/daglib/0090259,levasseur2012applied}]
  An element $a \ne 0$ of a Boolean algebra $\mathbf{B}$ is an \emph{atom} if
  there is no $x \in \mathbf{B}$ such that $0 < x < a$. Equivalently, $a \ne 0$
  is an atom if, for all $x \in \mathbf{B}$, either $x \land a = a$ or $x \land
  a = 0$. A Boolean algebra is \emph{atomic} if for every $a \in \mathbf{B}
  \setminus \{0 \}$, there is an atom $x$ such that $x \le a$.
\end{definition}

@@

\begin{lemma}[\cite{givant2008introduction}]
  All finite Boolean algebras are atomic.
\end{lemma}

\begin{theorem}[\cite{ganesh2006introduction}] \label{thm:representation}
  Let $\mathbf{B}$ be a finite Boolean algebra. Then every $a \in \mathbf{B}
  \setminus \{ 0 \}$ can be uniquely expressed as $a = \bigvee_{i \in I} m_i$
  for some set of atoms $\{m_i\}_{i \in I}$.
\end{theorem}

% TODO: remove the requirement for being strictly positive
\begin{definition}[\cite{gaifman1964concerning}] \label{def:measure}
  A \emph{(strictly positive) measure} on a Boolean algebra $\mathbf{B}$ is a
  function $m : \mathbf{B} \to [0, 1]$ such that:
  \begin{enumerate}
  \item $m(1) = 1$, and $m(x) > 0$ for $x \ne
    0$; \label[property]{property:values}
  \item $m(x \lor y) = m(x) + m(y)$ for all $x, y \in \mathbf{B}$ whenever $x
    \land y = 0$. \label[property]{property:additive}
  \end{enumerate}
\end{definition}

@@

% TODO: weight can be zero
\begin{definition}
  Let $\mathbf{B}$ be a finite Boolean algebra, let $L$ be a subset of
  $\mathbf{B}$ such that every atom $m$ can be uniquely expressed as $m =
  \bigwedge_{i \in I} l_i$ for some $\{ l_i \}_{i \in I} \subseteq L$, and let
  $w : L \to \mathbb{R}_{>0}$ be arbitrary. The \emph{weighted model count}
  $\WMC : \mathbf{B} \to \mathbb{R}_{\ge 0}$ is defined as
  \[
    \WMC(a) = \begin{cases}
      0 & \text{if } a = 0 \\
      \prod_{i \in I} w(l_i) & \text{if } M \ni a = \bigwedge_{i \in I} l_i \\
      \sum_{i \in I} \WMC(m_i) & \text{if } \mathbf{B} \setminus (M \cup \{ 0
      \}) \ni a = \bigvee_{i \in I} m_i
    \end{cases}
  \]
  for any $a \in \mathbf{B}$. Furthermore, we define the \emph{normalised
    weighted model count} $\nWMC : \mathbf{B} \to [0, 1]$ as $\nWMC(a) =
  \frac{\WMC(a)}{\WMC(1)}$ for all $a \in \mathbf{B}$.
\end{definition}

@@

\begin{proposition}
  Let $\mathbf{B}$ be a finite Boolean algebra with a measure $m :
  \mathbf{B} \to [0, 1]$. Let $\mathbf{A}$ be a $\mathbf{B}$-valued functional
  polyadic algebra with domain $X$ and variables $I$. Let $m^* : \mathbf{A} \to
  \mathbb{R}_{\ge 0}$ be defined by
  \[
    m^*(p) = \sum_{\substack{\text{atoms }y \in \mathbf{B} \text{ s.t.}\\ \exists x \in X^I:\, y \le p(x)}} m(y)
  \]
  for all $p \in \mathbf{A}$. Then $m^*$ is a measure on $\mathbf{A}$.
\end{proposition}

@@

\begin{proof} % TODO: update the proof
  First, we can show that $m^*(1) = 1$ by observing that
  \[
    m^*(1) = \sum_{\text{atoms } y \in \mathbf{B}} m(y) = m \left(
      \bigvee_{\text{atoms } y \in \mathbf{B}} y \right) = m(1) = 1,
  \]
  where we use \cref{thm:representation} and express $1 \in \mathbf{B}$ as the
  supremum of all atoms in $\mathbf{B}$ \cite{ganesh2006introduction}. Clearly
  $m^*(p) \ge 0$ for all $p \in \mathbf{A}$, so we can restrict the codomain of
  $m^*$ to $[0, 1]$.

@@

@@

@@

% TODO: 1-2 examples of how a probability is calculated
% TODO: condition for a measurable BA to be representable as WMC
% TODO: can I show that a generalisation of WMC is the only possible probability measure (when it exists)?
% TODO: abstraction refinements as homomorphisms.
% TODO: definition of a measure-preserving homomorphism from Thomas Jech
% TODO: a Boolean algebra is approximable if its Stone space is approximable
% TODO: maybe the measure on BA should not be normalised?
% TODO: how do measures transfer to quotient algebras? Naturally!
% TODO: extension to infinite atomic Boolean algebras

>>>>

\documentclass{article}
\usepackage[utf8]{inputenc}
\usepackage[UKenglish]{babel}
\usepackage[UKenglish]{isodate}
\usepackage{fullpage}
\usepackage{amsthm}
\usepackage{amsfonts}
\usepackage{amsmath}
\usepackage{mathtools}
\usepackage[capitalise]{cleveref}
\usepackage{bm}
\usepackage{booktabs}
\usepackage[backgroundcolor=lightgray]{todonotes}

@@

\title{Weighted Model Counting as a Special Case of Polyadic Measure Algebras}
\author{Paulius Dilkas}

@@

\todo[inline]{Which definition do I actually need?}
\begin{definition}[\cite{DBLP:books/daglib/0090259,levasseur2012applied}]
  An element $a \ne 0$ of a Boolean algebra $\mathbf{B}$ is an \emph{atom} if
  there is no $x \in \mathbf{B}$ such that $0 < x < a$. Equivalently, $a \ne 0$
  is an atom if, for all $x \in \mathbf{B}$, either $x \land a = a$ or $x \land
  a = 0$. A Boolean algebra is \emph{atomic} if for every $a \in \mathbf{B}
  \setminus \{0 \}$, there is an atom $x$ such that $x \le a$.
\end{definition}

@@

\begin{lemma}[\cite{givant2008introduction}] \label{lemma:atomic}
  All finite Boolean algebras are atomic.
\end{lemma}

\begin{theorem} \label{thm:representation}
  Let $\mathbf{B}$ be a finite Boolean algebra. Then every $a \in \mathbf{B}
  \setminus \{ 0 \}$ can be uniquely expressed as $a = \bigvee_{i \in I} a_i$,
  where $\{ a_i \}_{i \in I}$ is the set of atoms in $\mathbf{B}$ such that $a_i
  \le a$ for all $i \in I$.
\end{theorem}
\begin{proof}
  A simple consequence of the theorem that every finite Boolean algebra is
  isomorphic to a field of subsets of a set, where the cardinality of the set is
  equal to the number of atoms in the Boolean algebra.
\end{proof}

\todo[inline]{Remove the requirement for being strictly positive}
\begin{definition}[\cite{gaifman1964concerning}] \label{def:measure}
  A \emph{(strictly positive) measure} on a Boolean algebra $\mathbf{B}$ is a
  function $m : \mathbf{B} \to [0, 1]$ such that:
  \begin{enumerate}
  \item $m(1) = 1$, and $m(x) > 0$ for $x \ne
    0$; \label[property]{property:values}
  \item $m(x \lor y) = m(x) + m(y)$ for all $x, y \in \mathbf{B}$ whenever $x
    \land y = 0$. \label[property]{property:additive}
  \end{enumerate}
\end{definition}

@@

\todo[inline]{Allow weight to be zero}
\begin{definition}
  Let $\mathbf{B}$ be a finite Boolean algebra, and let $M \subseteq \mathbf{B}$
  be its set of atoms. Let $L \subseteq \mathbf{B}$ be such that every atom $m
  \in M$ can be uniquely expressed as $m = \bigwedge_{i \in I} l_i$ for some $\{
  l_i \}_{i \in I} \subseteq L$, and let $w : L \to \mathbb{R}_{>0}$ be
  arbitrary. The \emph{weighted model count} $\WMC : \mathbf{B} \to
  \mathbb{R}_{\ge 0}$ is defined as
  \[
    \WMC(a) = \begin{cases}
      0 & \text{if } a = 0 \\
      \prod_{i \in I} w(l_i) & \text{if } M \ni a = \bigwedge_{i \in I} l_i \\
      \sum_{i \in I} \WMC(m_i) & \text{if } \mathbf{B} \setminus (M \cup \{ 0
      \}) \ni a = \bigvee_{i \in I} m_i
    \end{cases}
  \]
  for any $a \in \mathbf{B}$. Furthermore, we define the \emph{normalised
    weighted model count} $\nWMC : \mathbf{B} \to [0, 1]$ as $\nWMC(a) =
  \frac{\WMC(a)}{\WMC(1)}$ for all $a \in \mathbf{B}$.
\end{definition}

@@

\begin{proposition} \label{prop:polyadic_measure}
  Let $\mathbf{B}$ be a finite Boolean algebra with a measure $m :
  \mathbf{B} \to [0, 1]$. Let $\mathbf{A}$ be a $\mathbf{B}$-valued functional
  polyadic algebra with domain $X$ and variables $I$. Let $m^* : \mathbf{A} \to
  \mathbb{R}_{\ge 0}$ be defined by
  \[
    m^*(p) = \sum_{\substack{\text{atoms }y \in \mathbf{B} \text{ s.t.}\\ \exists x \in X^I:\, y \le p(x)}} m(y)
  \]
  for all $p \in \mathbf{A}$. Then $m^*$ is a measure on $\mathbf{A}$.
\end{proposition}

@@

\begin{proof}\todo[inline]{Update the proof w.r.t. definitions}
  First, we can show that $m^*(1) = 1$ by observing that
  \[
    m^*(1) = \sum_{\text{atoms } y \in \mathbf{B}} m(y) = m \left(
      \bigvee_{\text{atoms } y \in \mathbf{B}} y \right) = m(1) = 1,
  \]
  where we use \cref{thm:representation} and express $1 \in \mathbf{B}$ as the
  supremum of all atoms in $\mathbf{B}$ \cite{ganesh2006introduction}. Clearly
  $m^*(p) \ge 0$ for all $p \in \mathbf{A}$, so we can restrict the codomain of
  $m^*$ to $[0, 1]$.

@@

\begin{lemma} \label{lemma:simple_measure}
  Given the setup of \cref{prop:polyadic_measure} and $p \in \mathbf{A}$, if
  $p(x) = p(y)$ for all $x, y \in X^I$ (i.e., $p$ has no free variables), then
  \[
    m^*(p) = m(p(x))
  \]
  (for some $x \in X^I$) is an alternative (i.e., equivalent and simpler)
  definition of $m^*$.
\end{lemma}
\begin{proof}
  Fix some $x \in X^I$. Then
  \[
    m(p(x)) = m \left( \bigvee_{\substack{\text{atoms } y \in \mathbf{B} \text{
            s.t.}\\
          y \le p(x)}} y \right) = \sum_{\substack{\text{atoms } y \in \mathbf{B}
        \text{ s.t.}\\
        y \le p(x)}} m(y) =\sum_{\substack{\text{atoms } y \in \mathbf{B}
        \text{ s.t.}\\
        \exists x' \in X^I : y \le p(x')}} m(y) = m^*(p),
  \]
  where we use \cref{thm:representation} for the first step,
  \cref{def:measure} and \cref{lemma:atomic} for the second step, the
  assumptions of \cref{lemma:simple_measure} for the third step, and the
  definition of $m^*$ for the fourth one.
\end{proof}

@@

\subsection{New Results (an Example)}

@@

\begin{table}
  \centering
  \caption{Step-by-step derivation of how a more complex element of
    $\mathbf{A}$ acts on elements of $X^I$}
  \label{tbl:derivation}
  \begin{tabular}{ll}
    \toprule
    Element of $\mathbf{A}$ & Action on $X^I$ \\
    \midrule
    $p$ & $(x_1, x_2) \mapsto \mathsf{p}(x_1, x_2)$ \\
    $b(2)p$ & $(x_1, x_2) \mapsto \mathsf{p}(x_1, b)$ \\
    $\neg b(2)p$ & $(x_1, x_2) \mapsto \neg\mathsf{p}(x_1, b)$ \\
    $\bm\exists(1)\neg b(2)p$ & $(x_1, x_2) \mapsto \neg\mathsf{p}(a, b) \lor \neg\mathsf{p}(b, b) = \neg(\mathsf{p}(a, b) \land \mathsf{p}(b, b))$ \\
    $\bm\forall(1)b(2)p = \neg\bm\exists(1)\neg b(2)p$ & $(x_1, x_2) \mapsto \neg\neg(\mathsf{p}(a, b) \land \mathsf{p}(b, b)) = \mathsf{p}(a, b) \land \mathsf{p}(b, b)$ \\
    \bottomrule
  \end{tabular}
\end{table}

\begin{table}
  \centering
  \caption{Atoms $y \in \mathbf{B}$ (and their measures) such that $y \le
    \mathsf{p}(a, b) \land \mathsf{p}(b, b)$}
  \label{tbl:atoms}
  \begin{tabular}{lc}
    \toprule
    Atom $y \in \mathbf{B}$ & $m(y)$ \\
    \midrule
    $\mathsf{p}(a, b) \land \mathsf{p}(b, b) \land \mathsf{p}(a, a) \land \mathsf{p}(b, a)$ & $1 \times 0.5 \times 0.5 \times 0 = 0$ \\
    $\mathsf{p}(a, b) \land \mathsf{p}(b, b) \land \neg\mathsf{p}(a, a) \land \mathsf{p}(b, a)$ & $1 \times 0.5 \times 0.5 \times 0 = 0$ \\
    $\mathsf{p}(a, b) \land \mathsf{p}(b, b) \land \mathsf{p}(a, a) \land \neg\mathsf{p}(b, a)$ & $1 \times 0.5 \times 0.5 \times 1 = 0.25$ \\
    $\mathsf{p}(a, b) \land \mathsf{p}(b, b) \land \neg\mathsf{p}(a, a) \land \neg\mathsf{p}(b, a)$ & $1 \times 0.5 \times 0.5 \times 1 = 0.25$ \\
    \bottomrule
  \end{tabular}
\end{table}

Let us calculate the probability $\Pr(\forall x_1 \in X, \mathsf{p}(x_1, b))$.
The same expression can be translated into the notation for our polyadic algebra
$\mathbf{A}$ as $m^*(\bm\forall(1)b(2)p)$. Recall that $\bm\forall(1)b(2)p =
\neg\bm\exists(1)\neg b(2)p$. The effect of this function on an arbitrary
element of $X^I$ is derived step-by-step in \cref{tbl:derivation}. Since the
resulting function is constant (i.e., the logical formula has no free
variables), \cref{lemma:simple_measure} tells us that
\[
  m^*(\bm\forall(1)b(2)p) = m(\mathsf{p}(a, b) \land \mathsf{p}(b, b)) = m
  \left( \bigvee_{\substack{\text{atoms } y \in \mathbf{B} \text{ s.t.}\\
        y \le \mathsf{p}(a, b) \land \mathsf{p}(b, b)}} y \right) = \sum
  _{\substack{\text{atoms } y \in \mathbf{B} \text{ s.t.}\\
      y \le \mathsf{p}(a, b) \land \mathsf{p}(b, b)}} m(y).
\]
The resulting sum is over four atoms; these atoms and their probabilities are
listed in \cref{tbl:atoms}. Thus, we get that
\[
  m^*(\bm\forall(1)b(2)p) = 0 + 0 + 0.25 + 0.25 = 0.5.
\]

\todo[inline,caption={}]{The big TODO list
  \begin{itemize}
  \item Condition for a measurable Boolean algebra to be representable as WMC.
    Note: if all atoms are in $L$, then WMC can define any distribution.
  \item Extension to infinite (atomic?) Boolean algebras.
  \item How many extra variable do you need to add to make any probability
    distribution representable using WMC?
  \item Abstraction refinements as homomorphisms.
  \item Definition of a measure-preserving homomorphism from Jech's set theory
    book.
  \item A Boolean algebra is approximable if its Stone space is approximable.
  \end{itemize}
}
